--- 
title: "Mortality Analysis"
author: "Pritam Biswas | Manas Dresswala | Prasham Sheth | Swarna Bharati Mantena"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
---

# Introduction

<h1>DO YOU KNOW?</h1> 
<p>
The death rate is 7.7 deaths/1,000 population (2018 est.) on average across the world. This rate results in about 108 worldwide deaths per minute or 1.8 deaths every second. \

Bulgaria has the highest death rate of 15.3 deaths/1,000 population and Qatar has the least death rate of 1.53 deaths/1,000 population. 
</p>
<h2>Does this little information bring up questions in your head?</h2>

<h2>YES?</h2> 
<p>
It did for us too. These are some of the questions that we came up with: \
What about the other countries? How many people die there on average? \
How does the number of deaths vary across different countries?
Do we see any pattern among deaths between developed and under-developed/developing countries? 
What are the main causes of death? What are the statistics corresponding to this? \
What about child mortality? How are the statistics in case of childeren (aged less than 5 years)? 
How do the statistics vary in different age groups and on sex? 

With a bunch of these questions that are quite interesting, we decided that this would be our project.
</p>
<h1>WHAT DO WE WANT TO DO?</h1> 

We picked the three most interesting questions that struck our minds when we started exploring the data on this topic.

<h2>The three interesting questions!</h2> 
<p>
a) Understand the child mortality accross the countries in the world. Is the child mortality count coming down? If yes, why did that happen? \

b) We know that in today’s world there has been a tremendous advancement in healthcare. Does the life expectancy of a human increase now? Did the number of deaths actually come down because of that? Or is there any other reason for the mortality rate of a country? \

c) Lastly, we would like to understand the mortality rates of people of different ages. Is there some difference in the causes of death? Are there any patterns here? Are there some differences in the causes of death when compared location wise? \

Let us now get to other sections of this book that will help you understand better how we proceeded on working on this project.
</p>

<!--chapter:end:index.Rmd-->

# Data and Information sources
<p>
We gathered data from multiple sources as we tried to answer a variety of questions in this project.
Some of the data sources that we would be using are listed below - 
\
https://www.worlddata.info/life-expectancy.php\
http://ghdx.healthdata.org/gbd-results-tool \
http://ucdp.uu.se/downloads/ \
https://www.prio.org/Data/Armed-Conflict/Battle-Deaths/The-Battle-Deaths-Dataset-version-30/ \
https://www.mortality.org \
https://population.un.org/wpp/Download/Standard/Mortality/ \
\
We tried to understand mortality not only through time but also across different countries. As we collected the data from different sources, they are in different formats. To overcome this, we performed some data cleaning and set the data collected in a proper format. 
In addition to this, we did not find consistent data, meaning if through one source we get data for 10 years then in the other source we are getting data for 8 years. Data cleaning helped us overcome this problem too. 
</p>
\ \ \ \
<p>
A few facts and data mentioned in this report is from the following sources. \
Information Resources: \
https://www.cia.gov/library/publications/the-world-factbook/fields/346.html \
https://www.worldatlas.com/articles/measuring-conflict-and-disease-the-top-global-mortality-rates.html \
</p>

<!--chapter:end:02-data.Rmd-->

# Data transformation

<!--chapter:end:03-cleaning.Rmd-->

# Missing values

<!--chapter:end:04-missing.Rmd-->

# Cause vs (Age & Sex)


```{r , include=FALSE}

library(ggplot2)
library(rlist)
library(anchors)
library(plyr)
library(dplyr)
library(plotly)
library(ggmosaic)

```
<p>
To understand the mortality rates of the people of different age groups, we downloaded datasets from the website http://ghdx.healthdata.org/gbd-results-tool and then plotted the basic plots relating the number of deaths to different age groups and sex. \
As the ages were grouped this way in the actual dataset source, we preferred to take it this way. There is no particular reason for choosing unequal intervals of age. \
The plots and information stated corresponds to data collectced from the year 1990 to 2016. Also, wherever average is mentioned, it is an average taken per annum from 1990 to 2017. 
</p>

<h3>Reading data and forming dataframes</h3>
```{r}
###CODE TO READ DATA FROM THE FILES
#reading csv files into a dataframes and
#preprocessing, forming proper datasets for plotting---------------------
csv <- read.csv(file = 'total_all_causes.csv', header = TRUE)
data_all <- as.data.frame(csv)


#preprocessing and forming proper datasets for plotting------------------

#--------------------------
#over all data
data_all_2 <- data_all[which(data_all$metric == 'Number'), ]
data_all_2$age <- plyr::revalue(data_all_2$age, c("1 to 4"="1-4 years"))
death_numbers_2 <- data_all_2 %>%
                  group_by(age, sex) %>%
                  dplyr::summarise(val_sum=sum(val),
                                   val_mean=(mean(val)))
death_numbers_2$age <- factor(death_numbers_2$age, levels = c("<1 year", "1-4 years", "5-14 years", "15-49 years", "50-69 years", "70+ years"))

#--------------------------
#injuries
csv3 <- read.csv(file = 'injuries.csv', header = TRUE)
data_injuries <- as.data.frame(csv3)
data_injuries2 <- data_injuries[which(data_injuries$metric == 'Number'), ]

data_injuries2$age <- plyr::revalue(data_injuries2$age, c("1 to 4"="1-4 years"))

data_injuries3 <- data_injuries2 %>%
                  group_by(age, cause) %>%
                  dplyr::summarise(val_sum=sum(val),
                                   val_mean=(mean(val)))
data_injuries3$age <- factor(data_injuries3$age, levels = c("<1 year", "1-4 years", "5-14 years", "15-49 years", "50-69 years", "70+ years"))

#---------------------------
#Deatsh due to communicable diseases
csv4 <- read.csv(file = 'communicable-1.csv', header = TRUE)
csv5 <- read.csv(file = 'communicable-2.csv', header = TRUE)
df4 <- as.data.frame(csv4)
df5 <- as.data.frame(csv5)
df4 <- df4[which(df4$metric == 'Number'), ]
df5 <- df5[which(df5$metric == 'Number'), ]
df_comm <- rbind(df4, df5)

df_comm$age <- plyr::revalue(df_comm$age, c("1 to 4"="1-4 years"))

data_comm_a <- df_comm %>%
                  group_by(age, cause) %>%
                  dplyr::summarise(val_sum=sum(val),
                                   val_mean=(mean(val)))
data_comm_a$age <- factor(data_comm_a$age, levels = c("<1 year", "1-4 years", "5-14 years", "15-49 years", "50-69 years", "70+ years"))

#---------------------------
#Deatsh due to communicable diseases
csv6 <- read.csv(file = 'non_communicable-1.csv', header = TRUE)
csv7 <- read.csv(file = 'non_communicable-2.csv', header = TRUE)
csv8 <- read.csv(file = 'non_communicable-3.csv', header = TRUE)
df6 <- as.data.frame(csv6)
df6 <- df6[which(df6$metric == 'Number'), ]
df7 <- as.data.frame(csv7)
df7 <- df7[which(df7$metric == 'Number'), ]
df8 <- as.data.frame(csv8)
df8 <- df8[which(df8$metric == 'Number'), ]
df_noncomm <- rbind(df6, df7, df8)

df_noncomm$age <- plyr::revalue(df_noncomm$age, c("1 to 4"="1-4 years"))

data_noncomm_a <- df_noncomm %>%
                  group_by(age, cause) %>%
                  dplyr::summarise(val_sum=sum(val),
                                   val_mean=(mean(val)))
data_noncomm_a$age <- factor(data_noncomm_a$age, levels = c("<1 year", "1-4 years", "5-14 years", "15-49 years", "50-69 years", "70+ years"))


```

<h3>Lets start plotting!</h3>


```{r}
#plot --- avg deaths vs age (all causes)
death_numbers_2a = aggregate(val_mean ~ age, data=death_numbers_2, FUN=sum)

plot_ly(x = death_numbers_2a$age, y = death_numbers_2a$val_mean, type = "bar") %>% 
  layout(title = "Avg. no. of deaths vs Age group",
         xaxis = list(title = "Age Groups",
                      zeroline = FALSE),
         yaxis = list(title = "Avg. no. of deaths",
                      zeroline = FALSE),
         font=list(family="Times New Roman", size=12))

```

<p>
From the plot, we see that the average number of deaths in the age group 70+ years is maximum and the average number of deaths in the age group 5-14 years is minimum. Also, it is interesting to see that the average number of deaths in the age groups 1-4 and 5-14 are less that the average number of deaths in age groups <1 year and 15-49 years. Overall, the average number of deaths decrease till 5-14 years age group and then increases. \
Why are the values as shown? What are the major causes of death? How do deaths due to various causes vary with age? These are the questions we wish to answer on further exploring and visualizing the dataset. \
</p>

<p>
Let us take a look at the causes of deaths (mentioned in the dataset).
</p>

```{r}

#Plot showing various causes of death
df_big <- rbind(data_injuries3, data_comm_a, data_noncomm_a)
df_big = aggregate(val_mean ~ cause, data=df_big, FUN=sum)

#df_big$cause <- factor(data$cause, levels = unique(data$cause)[order(data$val_mean, decreasing = TRUE)])
df_big <- df_big[order(df_big$val_mean),]

plot_ly(df_big, x = ~val_mean, y = ~cause, type = "bar")  %>% 
        layout(title="Causes of death vs Average nunber of deaths", 
               xaxis = list(title = "Average number of deaths",
                            zeroline = FALSE, family = "times new roman"),
               yaxis = list(title = "Causes of death",
                            zeroline = FALSE, family = "times new roman",
                            categoryorder = "array", categoryarray = df_big$cause),
               font=list(family="Times New Roman", size=12))

```
<p>
This plot shows the causes of deaths vs average number of deaths. As we see, mental disorders cause the least number of deaths while cardiovascular diseases cause the maximum number of deaths. \
The causes of death can be divided into three categories as follows: 1. injuries, 2. Communicable, maternal, neonatal, and nutritional diseases, and 3. Non-communicable diseases. Let us compare the number of deaths caused by each of these categories for the age groups.\

In this analysis, let us first check the deaths due to injuries of each age group. 
</p>

<h3>Injuries</h3>
```{r}

injuries_death_numbers = aggregate(val_mean ~ age, data=data_injuries3, FUN=sum)

plot_ly(x = injuries_death_numbers$age, y = injuries_death_numbers$val_mean, type = "bar") %>% 
        layout(title = "Avg. no. of deaths due to injuries vs Age group",
               xaxis = list(title = "Age Groups",
                            zeroline = FALSE),
               yaxis = list(title = "Avg. no. of deaths due to injuries",
                            zeroline = FALSE),
               font=list(family="Times New Roman", size=12))

```
<p>
From this graph, we can see that more number of deaths due to injuries are in the 15-49 years age group. It seems a little natural that this number would be high because the width of this age group is high. Now, let us dive into what the specific causes were, under this subgroup of injuries.
</p>
```{r}

plot_ly(data_injuries3, x = ~age, y = ~val_mean, color = ~cause, type = "bar")  %>% 
        layout(title="Avg. no. of deaths due to injuries vs Age", 
               xaxis = list(title = "Age Groups",
                            zeroline = FALSE),
               yaxis = list(title = "Avg. no. of deaths due to injuries",
                            zeroline = FALSE),
               font=list(family="Times New Roman", size=12))
    
```
<p>
This plot reveals a lot of information. Check the first three age groups. The people in these age groups and this category of causes of death (injuries) mostly died because of unintentional injuries. The common unintensional injuries include drowning, falls, poisoning, etc. As kids do not know how to handle the situations at their age, there are more chances of them losing life due to unintentional injuries as shown in the plot. Also, self-harm and interpersonal violence are very less in these three age groups. I think one of the most important reason for that is that people usually remain carefree during this age. They usually do not have any real life worries in life. \
In contrast, the number of deaths due to self-harm and interpersonal violence is most in the age group of 15-49 years (middle-age group). The stress levels are pretty high in this age group. The reasons can be many ranging from financial problems to heart breaks. \
In all the age groups, the average number of deaths due to self-harm and interpersoanl violence is less than the other two causes except for the age group 15-49 years. \

Following is a mosaic plot that gives the information of the same data in relative fashion. Here, we do not know the actual values by observing the plot alone. So, bar graph using plotly seem to be a better option than mosaic plots.  
</p>
```{r, fig.width=12, fig.height=8}

data_injuries4 <- data_injuries3 %>% select(age, cause, val_mean)
data_injuries4$val_mean <- as.integer(data_injuries4$val_mean) 
colnames(data_injuries4) <- c("age", "cause", "Freq")

ggplot(data_injuries4) + 
  geom_mosaic(
    aes(x=product(cause, age), # cut from right to left 
        weight=Freq,
        fill=cause
        ),
    divider=c("vspine" , "hspine") # equivalent to divider=ddecker()
  )  +
  ggtitle("Mosaic plot") +
  xlab("Age groups") + ylab("Causes (in injuries category)")

```



<h3>Communicable, maternal, neonatal, and nutritional diseases</h3>
```{r}

communicable_death_numbers = aggregate(val_mean ~ age, data=data_comm_a, FUN=sum)

plot_ly(x = communicable_death_numbers$age, y = communicable_death_numbers$val_mean, type = "bar") %>% 
        layout(title = "Avg. no. of deaths due to communicable diseases vs Age group",
               xaxis = list(title = "Age Groups",
                            zeroline = FALSE),
               yaxis = list(title = "Avg. no. of deaths due to communicable diseases",
                            zeroline = FALSE),
               font=list(family="Times New Roman", size=12))

```

<p>
From this graph, we can see that maximum average number of deaths in this category are in the age group of <1 year and the minimum average number of deaths are in the age group 4-15 years. Let us explore further on what this category of causes includes and how each of the causes is responsible for deaths. 
</p>

```{r}

plot_ly(data_comm_a, x = ~age, y = ~val_mean, color = ~cause, type = "bar")  %>% 
        layout(title="Avg. no. of deaths due to communicable diseases vs Age", 
               xaxis = list(title = "Age Groups",
                            zeroline = FALSE, family = "times new roman"),
               yaxis = list(title = "Avg. no. of deaths communicable diseases",
                            zeroline = FALSE, family = "times new roman"),
               font=list(family="Times New Roman", size=12))
    
```

<p>
In the age group <1 year, the one cause that is responsible for maximum number of deaths is 'Maternal and neonatal disorders'. This cause is prominnant only inn two age groups <1 year and 15-49 years. In all other age groups, the average number of deaths due to this is negligible. The age group 15-49 has the maximum average number of deaths dur to 'HIV/AIDS and sexually transmitted infections'. In all the age groups, 'Respiratory infections and tuberculosis' seems to be one the main reasons of deaths under this category causes of deaths. \

Again, following is the mosaic plot that gives us the information about the data in relative sense. Similar to the previous mosaic plot, we do not get to know the true values from mosaic plots. Also, as the set of boxes representing same cause do not start at the same point, it gets difficult to compare too. In addition, it also gets difficult as the number of causes increase. So, we will no longer check the mosaic plots for our graphs when the number of causes are high.
</p>

```{r, fig.width=12, fig.height=8}

data_comm_b <- data_comm_a %>% select(age, cause, val_mean)
data_comm_b$val_mean <- as.integer(data_comm_b$val_mean) 
colnames(data_comm_b) <- c("age", "cause", "Freq")

ggplot(data_comm_b) + 
  geom_mosaic(
    aes(x=product(cause, age), # cut from right to left 
        weight=Freq,
        fill=cause
        ),
    divider=c("vspine" , "hspine") # equivalent to divider=ddecker()
  ) +
  ggtitle("Mosaic plot") +
  xlab("Age groups") + ylab("Causes \n (in Communicable, maternal, neonatal, and nutritional diseases category)")

```


<h3>Non Communicable diseases</h3>
```{r}

non_communicable_death_numbers = aggregate(val_mean ~ age, data=data_noncomm_a, FUN=sum)

plot_ly(x = non_communicable_death_numbers$age, y = non_communicable_death_numbers$val_mean, type = "bar") %>% 
        layout(title = "Avg. no. of deaths due to non-communicable diseases vs Age group",
               xaxis = list(title = "Age Groups",
                            zeroline = FALSE),
               yaxis = list(title = "Avg. no. of deaths due to non-communicable diseases",
                            zeroline = FALSE),
               font=list(family="Times New Roman", size=12))

```

<p>
The trend in the plot is similar to the trend in the first plot. The average number of deaths decrease till the 5-14 years age group and then increases. The increase observed is very steep. The average number of deaths in the age group 70+ years is more than the sum of the average number of deaths of all other age groups under this category of causes of deaths. Let us explore further on what were the main causes under this category. 
</p>

```{r}

plot_ly(data_noncomm_a, x = ~age, y = ~val_mean, color = ~cause, type = "bar")  %>% 
        layout(title="Avg. no. of deaths due to non-communicable diseases vs Age", 
               xaxis = list(title = "Age Groups",
                            zeroline = FALSE, family = "times new roman"),
               yaxis = list(title = "Avg. no. of deaths non-communicable diseases",
                            zeroline = FALSE, family = "times new roman"),
               font=list(family="Times New Roman", size=12))
    
```

<p>
There seem to be less average number of deaths due to non-communicable diseases in the three age groups <1 year, 1-4 years, and 5-14 years. Cardiovascular diseases and Neoplasms are the two main non-communicale diseases that have recoded maximum average number of deaths in this category of causes of deaths.
</p>
<h3>Let us now check the average number of deaths and causes based on sex.</h3>

```{r}

plot_ly(death_numbers_2, x = ~age, y = ~val_mean, color = ~sex, type = "bar")  %>% 
  layout(title="Avg. no. of deaths vs Age (by sex)", 
         xaxis = list(title = "Age Groups",
                      zeroline = FALSE),
         yaxis = list(title = "Avg. no. of deaths",
                      zeroline = FALSE),
         font=list(family="Times New Roman", size=12))
    
```

<p>
Comparing the two graphs above, we see that irrespective of the sex, the same pattern of decrese in the average number of deaths till 5-14 years age group and then increase till 70+ years age group is observed.\

It can also be observed from the graph that the average number of deaths in males was greater than the average number of deaths in females under all age groups considered except '70+ years' group. Let us explore the division on the average number of death based on sex.
</p>

```{r}

death_numbers_2 <- death_numbers_2 %>% select(age, sex, val_mean)
death_numbers_2$val_mean <- as.integer(death_numbers_2$val_mean) 
colnames(death_numbers_2) <- c("age", "sex", "Freq")

ggplot(death_numbers_2) + 
  geom_mosaic(
    aes(x=product(sex, age), # cut from right to left 
        weight=Freq,
        fill=sex
        ),
    divider=c("vspine" , "hspine") # equivalent to divider=ddecker()
  ) +
  ggtitle("Mosaic plot") +
  xlab("Age groups") + ylab("Sex")

```

<p>
Here we come back to mosaic plots as it would be easy for comparisons. By looking at the graph we can easily compare the relative average number of deaths in males and females. But again, the accurate value is not known using mosaic plots. 
</p>

```{r}

#DATA
data_injuries_sex <- data_injuries2 %>%
                  group_by(sex, cause) %>%
                  dplyr::summarise(val_sum=sum(val),
                                   val_mean=(mean(val)))

data_comm_sex <- df_comm %>%
                  group_by(sex, cause) %>%
                  dplyr::summarise(val_sum=sum(val),
                                   val_mean=(mean(val)))

data_noncomm_sex <- df_noncomm %>%
                  group_by(sex, cause) %>%
                  dplyr::summarise(val_sum=sum(val),
                                   val_mean=(mean(val)))

```

```{r}

#PLOT injuries
plot_ly(data_injuries_sex, x = ~sex, y = ~val_mean, color = ~cause, type = "bar")  %>% 
  layout(title="Avg. no. of deaths vs Sex (by Injuries)", 
           xaxis = list(title = "Sex",
                      zeroline = FALSE),
         yaxis = list(title = "Avg. no. of deaths",
                      zeroline = FALSE),
         font=list(family="Times New Roman", size=12))
    
```

<p>
This graph shows the average number of deaths of females and males by the causes of death under the injuries category. Overall, it can be observed that majority of the deaths were due to unintensional injuries. In all the three causes included in the graph, the average number of deaths of females is less than that of males. Deaths due to transport injuries are least among the three in case of females. 
</p>

```{r}

#PLOT comm...
plot_ly(data_comm_sex, x = ~sex, y = ~val_mean, color = ~cause, type = "bar")  %>% 
  layout(title="Avg. no. of deaths vs Sex \n (by communicable, maternal, neonatal, and nutritional diseases)", 
           xaxis = list(title = "Sex",
                      zeroline = FALSE),
         yaxis = list(title = "Avg. no. of deaths",
                      zeroline = FALSE),
         font=list(family="Times New Roman", size=12))
    
```

<p>
From this graph, we observe that the average number of deaths due to enteric infections, HIV/AIDS nad sexually transmitted infections, neglected tropical diseases and malaria, nutritional deficiencies, and other infectious diseases is almost equal in case of females and males. \
Huge variation is observed in the average number of deaths due of maternal and neonatal disorders and due to respiratory infections and tuberculosis. In both cases, the average is more in males than that in females. In females, respiratory infections and tuberculosis causes the maximum average number of deaths in this category of deaths and in males, maternal and neonatal disorders cause the maximum average number of deaths. 
</p>

```{r}

#PLOT non-communicable diseases
plot_ly(data_noncomm_sex, x = ~sex, y = ~val_mean, color = ~cause, type = "bar")  %>% 
  layout(title="Avg. no. of deaths vs Sex (by non-communicable diseases)", 
           xaxis = list(title = "Sex",
                      zeroline = FALSE),
         yaxis = list(title = "Avg. no. of deaths",
                      zeroline = FALSE),
         font=list(family="Times New Roman", size=12))
    
```

<p>
In the category of non-communicable diseases, the average number of deaths due to diabetes is approximately equal in both females and males. Cardiovascular diseases and neoplasms (cancer) are the top two cause of death in both males and females under this category of causes of death. Both these have higher count in males than in females. The only disease that caused significantly higher deaths in females than in males under this category of causes of death is neurological disorders. \
\
NOTE: We tried to include cleaveland dot plot / scatter plot, but felt that including bar charts is a better choice on comparing them because of the ease of comparison of values in the bar charts. 
</p>

<!--chapter:end:05-cause_age_sex.Rmd-->

# HALE

```{r, include=FALSE}

library(ggplot2)
library(plotly)
library(gganimate)
library(rworldmap)
library(choroplethr)
library(dplyr)
library(tidyverse)

```




One important method of assessing the health of a population is to ask how long people can expect to live. 


Life expectancy, estimate of the average number of additional years that a person of a given age can expect to live. The most common measure of life expectancy is life expectancy at birth. Life expectancy is a hypothetical measure. It assumes that the age-specific death rates for the year in question will apply throughout the lifetime of individuals born in that year. The estimate, in effect, projects the age-specific mortality (death) rates for a given period over the entire lifetime of the population born (or alive) during that time. The measure differs considerably by sex, age, race, and geographic location. Therefore, life expectancy is commonly given for specific categories, rather than for the population in general.

Life expectancy reflects local conditions. In less-developed countries, life expectancy at birth is relatively low, compared with more-developed countries. In some less-developed countries, life expectancy at birth may be lower than life expectancy at age 1, because of high infant mortality rates (commonly due to infectious disease or lack of access to a clean water supply).

Life expectancy is calculated by constructing a life table. A life table incorporates data on age-specific death rates for the population in question, which requires enumeration data for the number of people, and the number of deaths at each age for that population. Those numbers typically are derived from national census and vital statistics data, and from them the average life expectancy for each of the age groups within the population can be calculated.

A life table is a table which shows, for a person at each age, what the probability is of them dying before their next birthday. From this starting point, a number of statistics can be derived and are also included in the table:

- the probability of surviving any particular year of age 
- the remaining life expectancy for people at each age
- the proportion of the original birth cohort still alive.
Life tables are usually constructed separately for men and for women because of their substantially different mortality rates. The Office of National Statistics is one organisation which produces life tables.

Health expectancy measures such as health-adjusted life expectancy (HALE) combine life expectancy (LE) with a measure of health-related quality of life (HRQL) or disability to create an indicator for assessing the combined effects of health and mortality, which is expressed in an intuitive measure similar to that of life expectancy [3]. Furthermore, in populations where life expectancy is increasing, health expectancies can be used to monitor whether the proportion of life spent in health is increasing (compression of morbidity) or decreasing (expansion of morbidity) due to a particular health problem such as insufficient or excess body weight 

Health Adjusted Life Expectancy (HALE):

HALE is a measure of population health that takes into account mortality and morbidity. It adjusts overall life expectancy by the amount of time lived in less than perfect health. This is calculated by subtracting from the life expectancy a figure which is the number of years lived with disability multiplied by a weighting to represent the effect of the disability.

If:

A = years lived healthily
B = years lived with disability

A+B = life expectancy

A+fB = healthy life expectancy, where f is a weighting to reflect disability level.


Healthy life expectancy (HALE) is a form of health expectancy that applies disability weights to health states to compute the equivalent number of years of good health that a newborn can expect.

Overall, global HALE at birth in 2015 for males and females combined was 63.1 years, 8.3 years lower than total life expectancy at birth. In other words, poor health resulted in a loss of nearly 8 years of healthy life, on average globally. Global HALE at birth for females was only 3 years greater than that for males. In comparison, female life expectancy at birth was almost 5 years higher than that for males.

HALE at birth ranged from a low of 51.1 years for African males to 70.5 years for females in the WHO European Region. The equivalent “lost” healthy years (LHE = total life expectancy minus HALE) ranged from 13% of total life expectancy at birth in the WHO African Region to 10% the WHO Western Pacific Region


("`https://www.healthknowledge.org.uk/e-learning/health-information/population-health-specialists/lifetables-hales-dalys-pylls#:~:targetText=Health%20Adjusted%20Life%20Expectancy%20(HALE,in%20less%20than%20perfect%20health.`")

("https://www.britannica.com/science/life-expectancy")

("https://www.who.int/gho/mortality_burden_disease/life_tables/hale_text/en/")




```{r, echo=FALSE}
Data_country_years <- read.csv(file="IHME-GBD_2017_DATA-country.csv", header=TRUE, sep=",")
Data_2017 <- read.csv(file="IHME-GBD_2017_DATA-c482ed82-1.csv", header=TRUE, sep=",")
Data_global_years <- read.csv(file="IHME-GBD_2017_DATA-7725bb0c-1.csv", header=TRUE, sep=",")

```


Below is the plot for the HALE values for the year of 2017. Hover over the points to see the country name. The plot is ordered in the decreasing order the HALE values.

The best thing about the plot below is you can hover over the points to see the country names and also zoom in and out by drawing boxes! The plot has 195 so do take out the time to explore the values for as many as you could by zooming on various parts of the graph.
```{r, echo=FALSE}

f <- list(
  family = "Courier New, monospace",
  size = 18,
  color = "#7f7f7f"
)
y <- list(
  title = "Country",
  titlefont = f
)
x <- list(
  title = "HALE (in years)",
  titlefont = f
) 


plot_ly(Data_country_years[Data_country_years$year == 2017,], x = ~val, y = ~reorder(location,val)) %>% 
  add_markers()%>%
  layout(xaxis = x, yaxis = y)

```
After rank ordering the countries by the HALE values, we selected to create an animation for top 10 and last 10 countries for showing the transition intheir values over the period of years 1990 to 2017. 

As we can see from them Japan and Singapore are 2 the places which have maintained their top positions. Andora, Sweden and Iceland inspite of increasing their respective values for HALE drop down in the positions. We think, availibility of better medical facilities along with the development in technology and easy adaptation in other countries as compared to that in these countries might have been the cause for the same. Also, we can see that the bottom most countries are the ones that are still developing ones and have less technological support available for helping them out through the medical situations. Thus, we can say that technological development does play an important role for helping us humans to survive more. 
```{r, fig.show='animate', echo=FALSE}


my <- Data_country_years %>%
  group_by(year) %>%
  mutate(rank = rank(-val)) %>%
  group_by(location) %>% 
  filter(rank <=10) %>%
  ungroup()
  
  
my1 <- Data_country_years %>%
  group_by(year) %>%
   mutate(rank = rank(-val))  %>%
  group_by(location) %>% 
  filter(rank > 185) %>%
  ungroup()

my <- rbind(my,my1)
my <- my %>%
  group_by(year) %>%
  mutate(rank = rank(-val),
         Value_rel = val/val[rank==1],
         Value_lbl = paste0(" ",round(val/1e9)))

staticplot = ggplot(my, aes(rank, group = location, 
                fill = as.factor(location), color = as.factor(location))) +
  geom_tile(aes(y = val/2,
                height = val,
                width = 0.9), alpha = 0.8, color = NA) +
  geom_text(aes(y = 0, label = paste(location, " ")), vjust = 0.2, hjust = 1) +
  geom_text(aes(y=val,label = val, hjust=0)) +
  coord_flip(clip = "off", expand = FALSE) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_reverse() +
  labs (x = "Countries", y = "Life Expectancy in years") +
  guides(color = FALSE, fill = FALSE) +
  theme(axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_text(size=18, hjust=0.5, face="italic", color="Black"),
        axis.title.y=element_blank(),
        legend.position="none",
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.grid.major.x = element_line( size=.1, color="grey" ),
        panel.grid.minor.x = element_line( size=.1, color="grey" ),
        plot.title=element_text(size=25, hjust=0.5, face="bold", colour="grey", vjust=-1),
        plot.subtitle=element_text(size=18, hjust=0.5, face="italic", color="grey"),
        plot.caption =element_text(size=8, hjust=0.5, face="italic", color="grey"),
        plot.background=element_blank(),
       plot.margin = margin(2,2, 2, 4, "cm"))


anim = staticplot + transition_states(year, transition_length = 4, state_length = 1) +
  view_follow(fixed_x = TRUE)  +
  labs(title = 'GDP per Year : {closest_state}',  
       subtitle  =  "Top 10 Countries",
       caption  = "GDP in Billions USD | Data Source: World Bank Data")

anim = staticplot + transition_states(year, transition_length = 4, state_length = 1) +
  view_follow(fixed_x = TRUE)  +
  labs(title = 'HALE per Year : {closest_state}',  
       subtitle  =  "Top 10 Countries")

animate(anim, 200, fps = 5,  width = 1200, height = 1000, 
        renderer = ffmpeg_renderer())
```

The below graph extends the above one in the sense that it has 40 countries into consideration (Top20 and bottom 20 when ranked based on the HALE values) The still graph with these many countries helps us see the fact that there is wide gap of 15 years when shifting from the 20th country from top to 20th country from bottom. 15 years! A big gap it is for sure...
```{r, fig.height= 12, fig.width=14, echo=FALSE}
Data <- Data_country_years %>% filter(year== 2017)


my <- Data %>%
  mutate(rank = rank(-val)) %>%
  group_by(location) %>% 
  filter(rank <=20) %>%
  ungroup()
  
  
my1 <- Data %>%
   mutate(rank = rank(-val))  %>%
  group_by(location) %>% 
  filter(rank > 175) %>%
  ungroup()

my <- rbind(my,my1)

my <- my %>%
   mutate(rank2 = rank(-val))

staticplot = ggplot(my, aes(rank2, group = location, fill = as.factor(location), color = as.factor(location))) +
  geom_tile(aes(y = val/2,height = val,width = 0.9), alpha = 1, color = NA) +
  geom_text(aes(y = 0, label = paste(location, " ")),size = 3, vjust = 0.2, hjust = 1) +
  geom_text(aes(y=val,label = val, hjust=0),size = 3) +
  coord_flip(clip = "off", expand = FALSE) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_reverse() +
  labs (x = "Countries", y = "Life Expectancy in years") +
  guides(color = FALSE, fill = FALSE) +
  theme(axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_text(size=18, hjust=0.5, face="italic", color="Black"),
        axis.title.y=element_blank(),
        legend.position="none",
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.grid.major.x = element_line( size=.1, color="grey" ),
        panel.grid.minor.x = element_line( size=.1, color="grey" ),
        plot.title=element_text(size=25, hjust=0.5, face="bold", colour="grey", vjust=-1),
        plot.subtitle=element_text(size=18, hjust=0.5, face="italic", color="grey"),
        plot.caption =element_text(size=8, hjust=0.5, face="italic", color="grey"),
        plot.background=element_blank(),
       plot.margin = margin(2,2, 2, 4, "cm"))


staticplot
```
Now, let's shift gears towards analyzing the HALE values for different genders over the countries. Below is the plot showing similar information as the first one, but here we have seperated the values for different genders and as can be seen for almost each country there is a clear trend of females having more HALE as compared to males. STRANGE thing for such a spectacular pattern to exist over all the countries! 

We need to study the different aspects for understanding why the pattern exists and get the answer to the question "Why do women live longer than men?"

```{r, echo=FALSE}

plot_ly(Data_2017, x = ~val, y = ~reorder(location,val),
        color = ~sex) %>%
  add_markers()%>%
  layout(xaxis = x, yaxis = y)

```

From the below drawn line and barcharts also it is a clear trend that the HALE values for females have been higher than the ones for males since 1990s and the trend has remained the same over the period of years.
```{r, width=10, echo=FALSE}

ggplotly(ggplot(Data_global_years, aes(year, val, color=sex)) + 
    geom_line() +
    ggtitle("Health Adjusted Life Expectancy") +
    labs (x = "", y = "Life Expectancy in years") +
    theme_grey(14) +
    theme(legend.title = element_blank())+
    scale_x_continuous(name="Years", (breaks=seq(1990,2017,2))))


p<-ggplot() + geom_bar(data=Data_global_years, aes(x = year, y=val, fill=sex), stat="identity", position="dodge") + scale_x_continuous(name="Years", (breaks=seq(1990,2017,2))) + ggtitle("Health Adjusted Life Expectancy") +
    labs (x = "", y = "Life Expectancy in years") + theme(legend.position="bottom")
  #coord_cartesian(ylim=c(50,65))


p


```
Combining the trends 

```{r, echo=FALSE, message=FALSE, warning=FALSE, comment=FALSE, echo=FALSE}
g <- list(
  showframe = FALSE,
  showcoastlines = FALSE,
  projection = list(type = 'Mercator')
)


Data <- Data_country_years %>% filter(year== 2017)

Data2m <- joinCountryData2Map( Data, 
                               nameJoinColumn="location", 
                               joinCode="NAME" )
mapCountryData(Data2m, 
                nameColumnToPlot='val', 
                catMethod='fixedWidth', 
                numCats=100,
                mapTitle="Health Adjusted Life Expectancy in Years(2017): Both Sexes",
                missingCountryCol="white"
)


Data2female <- Data_2017 %>% filter(sex=="Female")

Data2m <- joinCountryData2Map( Data2female, 
                               nameJoinColumn="location", 
                               joinCode="NAME" )
mapCountryData(Data2m, 
                nameColumnToPlot='val', 
                catMethod='fixedWidth', 
                numCats=100,
                mapTitle="Health Adjusted Life Expectancy in Years(2017): Female",
                missingCountryCol="white"
)

Data2male <- Data_2017 %>% filter(sex=="Male")

Data2m <- joinCountryData2Map( Data2male, 
                               nameJoinColumn="location", 
                               joinCode="NAME" )
mapCountryData(Data2m, 
                nameColumnToPlot='val', 
                catMethod='fixedWidth', 
                numCats=100,
                mapTitle="Health Adjusted Life Expectancy in Years(2017): Male",
                missingCountryCol="white"
)


```



```{r echo=FALSE, fig.width=20, fig.height=20}

d1<- Data_country_years[Data_country_years$location == "United States" ,]
d2<- Data_country_years[Data_country_years$location == "India" ,]
d3<- Data_country_years[Data_country_years$location == "Canada" ,]
d4<- Data_country_years[Data_country_years$location == "Zimbabwe" ,]
d<-rbind(d1,d2,d3,d4)
ggplotly(ggplot(d, aes(year, val)) + 
    geom_line() +
    ggtitle("Health Adjusted Life Expectancy") +
    labs (x = "", y = "Life Expectancy in years") +
    theme_grey() +
    theme(legend.title = element_blank())+
  facet_wrap(~location,  ncol=2)+
  scale_x_continuous(name="Years", (breaks=seq(1990,2017,3))))


ggplotly(ggplot(d1, aes(year, val)) + 
    geom_line() +
    ggtitle("Health Adjusted Life Expectancy") +
    labs (x = "", y = "Life Expectancy in years") +
    theme_grey(14) +
    theme(legend.title = element_blank())+
  scale_x_continuous(name="Years", (breaks=seq(1990,2017,3))))
```



```{r ch1unk-label, include=FALSE , message=FALSE, warning=FALSE, comment=FALSE}

datalow <- Data_country_years %>% rename(region= location, value = val) %>% 
  mutate(region = tolower(region)) %>%
  mutate(region = recode(region,"united states"    = "united states of america", "russian federation"="russia","congo"= "republic of congo", "tanzania" = "united republic of tanzania", "serbia" = "republic of serbia"))

choropleths = list()
j=1

for (i in 1:nrow(datalow))
{
  if (is.na(datalow[i,"value"])) 
    {
    datalow[i,"breaks"] <- "NA"
  } else if (datalow[i,"value"] < 40) {
    datalow[i,"breaks"] <- "0-40"
  } else if (datalow[i,"value"]>=40 & datalow[i,"value"]<45) {
    datalow[i,"breaks"] <- "40-45"
  } else if (datalow[i,"value"]>=45 & datalow[i,"value"]<50) {
    datalow[i,"breaks"] <- "45-50"
  } else if (datalow[i,"value"]>=50 & datalow[i,"value"]<55) {
   datalow[i,"breaks"] <- "50-55"
  } else if (datalow[i,"value"]>=55 & datalow[i,"value"]<60) {
    datalow[i,"breaks"] <- "55-60"
  } else if (datalow[i,"value"]>=60 & datalow[i,"value"]<65) {
    datalow[i,"breaks"] <- "60-65"
  } else if (datalow[i,"value"]>=65 & datalow[i,"value"]<70) {
   datalow[i,"breaks"] <- "65-70"
  }else if (datalow[i,"value"]>=70 & datalow[i,"value"]<75) {
   datalow[i,"breaks"] <- "70-75"
  }else if (datalow[i,"value"]>=75) {
    datalow[i,"breaks"] <- "75+"
  }
}
datalow$value<-datalow$breaks


for (i in 1990:2017) {
  df           = datalow %>% filter(year== i)
  title        = paste0("Year: ",i)
  choropleths[[j]] = country_choropleth(df, title=title, num_colors = 5)
  j=j+1
}
choroplethr_animate(choropleths)


```

<iframe src="animated_choropleth.html" width="1200" height="2000"></iframe>


<!--chapter:end:06-Health_Adjusted_Life_Expectancy.Rmd-->

# Interactive component



<!--chapter:end:07-interactive.Rmd-->

